---
title: "Preventable Mortality Under 75: District-Level Variation in England"
author: "Gabriel Camacho-Reid"
date: "2011-2019"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(ggplot2)
library(ggrepel)

analytical_table <- readRDS("C:/Users/User/Documents/data/cleaned/analytical_table.rds")
gender_split <- read.csv("C:/Users/User/Documents/data/cleaned/query_result_gender.csv")

analytical_table <- analytical_table |> filter(AreaCode != "E09000001")
gender_split <- gender_split |> filter(areacode != "E09000001")
```

## Background

Preventable mortality — deaths from causes considered avoidable through effective public health intervention or healthcare — varies significantly across English districts. Understanding what drives this variation matters for how NHS resources are allocated and where public health intervention is most needed.

This analysis examines district-level variation in preventable mortality under 75 across England between 2011 and 2019, using data from NHS Fingertips, the ONS Index of Multiple Deprivation 2019, and the 2021 Census. It asks which districts outperform or underperform their deprivation profile, and what area-level characteristics distinguish them.

## Data and methods

Data on preventable mortality under 75 (NHS Fingertips indicator E03, IndicatorID 93721) was extracted for English districts and unitary authorities using the 2020/21 boundary definitions (AreaTypeID 301). Three-year rolling averages were used in preference to single-year figures to reduce statistical noise at district level. The analysis covers the period 2011–2019, excluding 2020 onwards to avoid the statistical discontinuity introduced by Covid-19 and the 2019 ONS revision to the preventable mortality definition.

Deprivation was measured using the IMD 2019 score aggregated to district level from LSOA-level data published by the Ministry of Housing, Communities and Local Government. Ethnic composition was derived from the 2021 Census (table TS021), expressed as the percentage of usual residents not identifying as White British. Gender breakdowns are built into the Fingertips indicator.

All data was loaded into a PostgreSQL database and joined on ONS area codes. Analysis was conducted in R. The City of London and Buckinghamshire UA were excluded due to anomalous population size and boundary mismatch respectively.

### Limitations

- Analysis is ecological — all associations are at area level, not individual level
- Ethnic minority composition reflects area characteristics, not individual outcomes by ethnicity
- Deprivation and ethnic minority composition are correlated in many districts, limiting independent interpretation
- Boundary changes across 2011–2019 affect comparability for some districts
- NHS reporting standards and cause of death coding changed across the study period
## Findings

### Deprivation is a strong predictor of preventable mortality — but not the whole story

Across English districts, IMD deprivation score is positively associated with preventable mortality under 75 for both men and women. Districts with the highest deprivation scores — Blackpool, Manchester, Liverpool, Middlesbrough — consistently show the highest mortality rates. Districts with the lowest deprivation — Hart, Wokingham, South Cambridgeshire — show the lowest.

However, deprivation alone does not fully explain the variation. A significant number of districts perform substantially better or worse than their deprivation score would predict.

```{r actual_vs_predicted, fig.cap="Figure 1: Actual vs predicted preventable mortality by district"}
district_avg <- analytical_table |>
  filter(Sex == "Persons" & Value != 0) |>
  group_by(AreaCode, AreaName) |>
  summarise(
    avg_mortality = mean(Value, na.rm = TRUE),
    avg_imd = mean(imd_score, na.rm = TRUE),
    avg_ethnicity = mean(pct_ethnic_minority, na.rm = TRUE),
    .groups = "drop"
  )

model <- lm(avg_mortality ~ avg_imd, data = district_avg)
district_avg$predicted <- predict(model, district_avg)
district_avg$residual <- district_avg$avg_mortality - district_avg$predicted

district_avg$label <- ifelse(
  abs(district_avg$residual) > 25 | district_avg$avg_ethnicity > 70,
  district_avg$AreaName, NA
)

ggplot(district_avg, aes(x = predicted, y = avg_mortality, colour = avg_ethnicity)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(alpha = 0.7, size = 2.5) +
  geom_label_repel(aes(label = label), size = 3, max.overlaps = 15,
                   fill = "white", label.padding = 0.15) +
  scale_colour_gradient(low = "#D3D1C7", high = "#534AB7",
                        name = "% ethnic minority") +
  labs(
    title = "Actual vs predicted preventable mortality by district",
    subtitle = "Points below the line outperform their deprivation profile",
    x = "Predicted mortality rate (from IMD)",
    y = "Actual average mortality rate (under 75)",
    caption = "Source: NHS Fingertips E03, ONS IMD 2019, Census 2021. England districts 2011–2019."
  ) +
  theme_minimal()
```

### High ethnic minority districts systematically outperform their deprivation profile

Districts with high ethnic minority populations — particularly London boroughs — consistently sit below the deprivation-predicted mortality line. Harrow, Enfield, Barnet, Newham, and Brent all show substantially lower mortality than their IMD scores would predict.

This pattern is concentrated in London but not exclusive to it. Birmingham, Bradford, and Leicester show similar though weaker effects. The consistency of the pattern across multiple London boroughs suggests a structural rather than coincidental explanation.

Two competing explanations are plausible. First, the healthy migrant effect — migrants are positively selected for health through income thresholds and physical fitness requirements, meaning high migration areas contain populations healthier than their area deprivation implies. Second, urban service concentration — London's density of NHS infrastructure, specialist services, and economic opportunity creates a protective effect that partially offsets deprivation penalties regardless of population composition.

These explanations are not mutually exclusive, and this analysis cannot distinguish between them at area level. Both are consistent with the observed pattern and both have implications for NHS resource allocation policy.

### The gender gap widens with deprivation

```{r gender_gap, fig.cap="Figure 2: Gender gap in preventable mortality by deprivation decile"}
district_gender <- analytical_table |>
  filter(Sex %in% c("Male", "Female") & Value != 0) |>
  mutate(imd_decile = ntile(imd_score, 10)) |>
  group_by(imd_decile, Sex) |>
  summarise(
    avg_mortality = mean(Value, na.rm = TRUE),
    .groups = "drop"
  )

gender_gap <- district_gender |>
  tidyr::pivot_wider(names_from = Sex, values_from = avg_mortality) |>
  mutate(gap = Male - Female)

ggplot(gender_gap, aes(x = imd_decile, y = gap)) +
  geom_col(fill = "#534AB7", alpha = 0.8) +
  geom_smooth(method = "lm", se = TRUE, colour = "#D85A30", linetype = "dashed") +
  scale_x_continuous(breaks = 1:10,
                     labels = c("1\nLeast\ndeprived", 2:9, "10\nMost\ndeprived")) +
  labs(
    title = "Gender gap in preventable mortality widens with deprivation",
    subtitle = "Difference between male and female mortality rates by deprivation decile, 2011–2019",
    x = "Deprivation decile (IMD 2019)",
    y = "Male minus female mortality rate",
    caption = "Source: NHS Fingertips E03, ONS IMD 2019. England districts 2011–2019."
  ) +
  theme_minimal()
```

Women have lower preventable mortality than men across all deprivation levels. However the gap is not constant — it widens substantially as deprivation increases, meaning men bear a disproportionately greater mortality burden in the most deprived areas.

### Ethnic minority composition predicts male outcomes more consistently than female

```{r gender_scatter, fig.cap="Figure 3: Preventable mortality by deprivation and gender"}
ggplot(gender_split, aes(x = avg_imd, y = avg_mortality, color = sex)) +
  geom_point(aes(size = avg_ethnicity), alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Female" = "#E05C8A", "Male" = "#378ADD")) +
  labs(
    title = "Preventable mortality by deprivation and gender",
    subtitle = "Bubble size = % ethnic minority population, 2011-2019",
    x = "Average IMD score",
    y = "Average mortality rate (under 75)",
    color = "Sex",
    size = "% ethnic minority",
    caption = "Source: NHS Fingertips E03, ONS IMD 2019, Census 2021. England districts 2011–2019."
  ) +
  theme_minimal()
```

The scatter of high ethnic minority districts around the female trend line is notably wider than around the male line. Ethnic minority composition is a stronger and more consistent predictor of male mortality outcomes than female outcomes.

This is consistent with gendered migration pathways. Male economic migrants — who disproportionately enter on skilled work visas with income thresholds — are more strongly health-selected than women who more commonly migrate through family reunion routes. The healthy migrant effect, to the extent it operates, is therefore likely a gendered phenomenon that applies more consistently to men than to women.
## Conclusions

Deprivation is the strongest single predictor of preventable mortality under 75 across English districts, but it is an incomplete explanation. Three findings complicate the simple deprivation narrative.

First, districts with high ethnic minority populations — concentrated in London but present elsewhere — consistently outperform their deprivation-predicted mortality rate. The most plausible explanations involve health selection effects among economic migrants and the protective effect of urban service concentration in London. This analysis cannot distinguish between these explanations at area level, but both carry implications for how the NHS interprets deprivation-based resource allocation in high migration urban areas.

Second, the mortality burden of deprivation falls disproportionately on men. The gender gap in preventable mortality widens as deprivation increases, suggesting that whatever protective factors exist in more deprived communities operate more strongly for women than for men.

Third, ethnic minority composition predicts male mortality more consistently than female mortality. This is consistent with gendered migration pathways — male economic migrants face stronger health selection through visa income thresholds than women who more commonly arrive through family reunion routes.

Taken together these findings suggest that area-level deprivation measures like IMD may systematically misrepresent health risk in high migration urban areas, and that the relationship between deprivation and mortality is moderated by both geography and gender in ways that a simple deprivation ranking does not capture.

## Future directions

Several avenues remain open for further analysis. A formal regression with an urban-rural classification and interaction terms between sex and geography would allow statistical testing of the urban service concentration hypothesis. Separating the ethnic minority composition variable by broad ethnic group — rather than treating it as a single percentage — would allow examination of whether the protective effect differs across communities. Extending the time series beyond 2019 with appropriate Covid adjustments would reveal whether the patterns observed here persisted through the recent period of NHS disruption.
